Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-06-22 appendix.tex
major rewrite to appendix for general branch lengths

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x,y,x,y,w\})$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$\\
    $\{1\}$     & $p_{1}$           &$1-x^2+y^2-x^2y^2$\\
    $\{2\}$     & $p_{2}$           &$1+x^2-y^2-x^2y^2$\\
    $\{3\}$     & $p_{3}$           &$1-x^2+y^2-x^2y^2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1+x^2-y^2-x^2y^2$\\
    $\{1,2\}$   & $p_{12}$          &$1-x^2-y^2+x^2y^2$\\
    $\{2,3\}$   & $p_{23}$          &$1-x^2-y^2+x^2y^2$\\
    $\{1,3\}$   & $p_{13}$          &$1+x^2+y^2-4xyw+x^2y^2$\\
    \hline
\end{tabular}
\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x,y,x,y,w\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

@@

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
Using branch fidelities, the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
Likelihood computations for all site splits and ancestral state splits are in Tables~\ref{tab:farris_likelihoods} (for the InvFels tree) and~\ref{tab:fels_likelihoods} (for the Felsenstein tree).
Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like00}
\caption[short]{}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like10}
\caption[short]{}
\end{subfigure}
\caption{
    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $x$, $y$, and $w$.
    Edges labeled by the probability of substitution along that edge.
    In (a), we compute the product to obtain $\Pr(\ancestralSplitRV=\emptyset\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
    In (b), the same process yields $\Pr(\ancestralSplitRV=\{1\}\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1-w)/32$.
}
\label{fig:example_likelihoods}
\end{figure}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$          &$(1-x)^2   (1-w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$       &$(1-x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)^2   (1-w)(1+y)^2$     &$(1+x)^2   (1+w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)^2   (1-w)(1+y)(1-y)$  &$(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
All values multiplied by $1/32$.}
\label{tab:farris_likelihoods}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_2, t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1-x)^2   (1-w)(1+y)(1-y)$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1-x)^2   (1-w)(1-y)^2$\\
     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)^2   (1-w)(1+y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1+x)^2   (1+w)(1-y)^2$\\
     $\{2,3\}$     &$(1-x)^2   (1-w)(1+y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1+x)^2   (1+w)(1+y)(1-y)$\\
\hline
\end{tabular}
\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the Felsenstein tree $\tau_2$.
All values multiplied by $1/32$.}
\label{tab:fels_likelihoods}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{InvFels tree ($\tau=\tau_1$)}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$&
    $\emptyset$&
    $(1+x)^2   (1+w)(1+y)^2$\\
     $\{1\}$    &
    $\emptyset$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
     $\{2\}$    &
    $\emptyset$&
    $(1+x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$    &
    $\emptyset$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
    $\{1,2\}$  &
    $\{\emptyset,\{1,2\}\}$&
    $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,3\}$  &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x)^2   (1+w)(1+y)^2\\
                    (1+x)^2   (1-w)(1+y)^2\\
                    (1+x)^2   (1+w)(1-y)^2
                \end{array}$\\
    $\{2,3\}$  &
                $\{\emptyset,\{1,2\}\}$&
                $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,2,3\}$&
                $\{1,2\}$&
                $(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
    \multicolumn{3}{c}{Felsenstein tree ($\tau=\tau_2$)}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$       &$\emptyset$&$(1+x)^2   (1+w)(1+y)^2$\\
    $\{1\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                        (1+x)(1-x)(1+w)(1+y)^2\\
                        (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
      $\{2\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2
                    \end{array}$\\
      $\{3\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)^2\\
                    (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
     $\{1,2\}$         &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1+x)^2   (1-w)(1+y)^2
                    \end{array}$\\
     $\{1,3\}$         &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{\{1\},\{2\}\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x)^2   (1+w)(1+y)^2\\
                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
                    (1+x)^2   (1+w)(1-y)^2
                    \end{array}$\\
      $\{2,3\}$        &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1-x)^2   (1-w)(1+y)^2\\
                    (1+x)^2   (1-w)(1-y)^2
                    \end{array}$\\
     $\{1,2,3\}$       &
    $\left\{\begin{array}{l}
                    \{2\} \\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1-w)(1+y)^2 \\
                    (1+x)^2   (1+w)(1+y)(1-y)
                    \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states on the InvFels tree $\tau_1$ and the Felsenstein tree $\tau_2$.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
See Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} for full calculations.}
\label{tab:likelihoods}
\end{table}
%TODO: the caption of this overlaps with the page number---will this be a problem or will MBE handle it?
%EM There is a latex package for that, which I tried out and then removed for some reason. It's no bigs.

\subsection*{Identifiability criteria and form of the likelihood}

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
In order for our maximum likelihood problem to be well-posed, we require identifiability conditions on the branch parameters.
We show that, in the case of the InvFels tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference yields an identifiable maximum likelihood inference procedure.

We first show that the existence of a unique maximizer implies $x_1=x_2$ and $y_1=y_2$.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

@@

\topoInconsist*

\begin{proof}
Assume $\tau^*=\tau_1$.
The InvFels tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).

We write out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.
Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is, suppressing normalizing constants $-\log 8$ (from the first additive term of \eqref{eq:site_pattern_profile_likelihood_mean}) and $-\log 32$ (from the second),

>>>>

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x,y,x,y,y\})$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xy^2+x^2y^2$\\
    $\{1\}$     & $p_{1}$           &$1-x^2+y^2-x^2y^2$\\
    $\{2\}$     & $p_{2}$           &$1+x^2-y^2-x^2y^2$\\
    $\{3\}$     & $p_{3}$           &$1-x^2+y^2-x^2y^2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1+x^2-y^2-x^2y^2$\\
    $\{1,2\}$   & $p_{12}$          &$1-x^2-y^2+x^2y^2$\\
    $\{2,3\}$   & $p_{23}$          &$1-x^2-y^2+x^2y^2$\\
    $\{1,3\}$   & $p_{13}$          &$1+x^2+y^2-4xy^2+x^2y^2$\\
    \hline
\end{tabular}
\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x,y,x,y,y\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

@@

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
Using branch fidelities, the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
Likelihood computations for all site splits and ancestral state splits are in Table~\ref{tab:farris_likelihoods} for the InvFels tree.
Taking maxima over ancestral state splits results in Table~\ref{tab:likelihoods}.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like00}
\caption[short]{}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like10}
\caption[short]{}
\end{subfigure}
\caption{
    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $\{x_1,y_1,x_2,y_2,w\}$.
    Edges labeled by the probability of substitution along that edge.
    In (a), we compute the product to obtain $\Pr(\ancestralSplitRV=\emptyset\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1+w)/32$.
    In (b), the same process yields $\Pr(\ancestralSplitRV=\{1\}\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1-w)/32$.
}
\label{fig:example_likelihoods}
\end{figure}
%TODO: update this figure with general branch lengths

\begin{table}
\centering
\begin{tabular}{|ll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplit_k$ & $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$\\
\hline
$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$\\
&$\{2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$\\

$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$\\
&$\{2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$\\

$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$\\
&$\{2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$\\

$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{2\}$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\

$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{2\}$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\

$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$\\
&$\{2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$\\

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{2\}$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\

$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
All values multiplied by $1/32$.}
\label{tab:farris_likelihoods}
\end{table}

\begin{table}
\centering
\begin{tabular}{|ll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
\hline
$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\

$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\

$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\

$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\

$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\

$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\

$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
Due to the symmetry of the likelihood, WLOG we assume $x_2 > x_1$ and $y_2 > y_1$ and maximize over ancestral states to reduce the number of possible functional forms to consider.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
See Table~\ref{tab:farris_likelihoods} for full calculations.}
\label{tab:likelihoods}
\end{table}
%TODO: the caption of this overlaps with the page number---will this be a problem or will MBE handle it?
%EM There is a latex package for that, which I tried out and then removed for some reason. It's no bigs.

\subsection*{Properties of the joint objective function}

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
We show that the likelihood remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

@@

Tentative theorem 1: 
\begin{restatable}{theorem}{topoInconsist2}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$.
There exists a set of $0 < x^*, y^* < 1$ such that
\[
\{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\} = \arg\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t)
\]
and $\hat{w}\equiv 1$.
\end{restatable}

In words, this implies an inconsistency as we converge on the incorrect topology, estimating $\hat{w}$ to be $1$ and not $y^*$.
